Regression Cube Analysis of Cohort Study Data

### 2015-03-29 draft.tex

% -Dependent & -Based in Überschriften groß, sicherheitshalber nochmal Steffi fragen
% oder: Dimension Reduction Depending on the Target Variable
\subsection{Target-Variable-Dependent Dimension Reduction}
% \com{Uli, Can you please proofread this?}
% %%
% The vast majority of features in an epidemiological data have no or a very low relation towards the target feature.
% %%
% Identifying these features und excluding them from the calculation can reduce the number of dimension significantly.
% %%
% The \emph{Correlation based Feature Selection} (CFS) algorithm is very popular in data mining for this task \cite{CFS}.
% %%
% It uses information entropy to select the features which have the most explanatory power towards the target feature.
% %%
% At the same time it tries to reduce the correlation between the independent features.
% %%
In epidemiological studies, manifold recordings lead to an abundance of features and thus a high-dimensional feature space.
%%
In general, many of them exhibit low or none correlation at all towards the target feature.
%%
Identifying irrelevant features and excluding them from the feature space considerably reduces computational costs and yields a comprehensible \emph{Regression Cube} representation.
%%
The correlation-based feature selection (CFS) \cite{CFS} aims to find a feature subset that maximizes the \emph{merit value} $M_F$ which is the ratio between the average feature-class and feature-feature dependencies of features in a given feature set $F$.
%%
The dependency of a set of features utilizes the entropy-based information gain to measure the explanatory power towards the target feature.
%%
Starting with an empty set of features $F$, the CFS algorithm iteratively adds the feature $f$ to $F$ that leads to the highest new merit-value $M_{F\cup f}$ and halts when no feature is left that would increase the merit.
%%
For example, if the \emph{body weight} has a strong explanatory power towards the target, it is likely that \emph{BMI} or \emph{waist circumference} exhibit similar correlations to the target.
%%
However, they strongly correlate with each other.
%%
The CFS algorithm will select the feature which has the largest explanatory power and discards the other dimensions.

@@

\subsection{Abstracting Regression Results Using $R^2$}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.5in]{figures/cube}
 \caption{
 (a) Overview visualization using a mosaic plot for the formula $Z \sim X + Y$, where $Z$ assumes the feature $Age$.
 %%
 The $R^2$ values extracted from the regression formulas depict the quality-of-fit and are mapped to color saturation (a saturated color shows a strong correlation).
 %%
 (b) Now, $Z$ is set to all features $n$ and yields $n$ mosaic plot visualizations.
 %%
 These represent the slices in our cube.
 %%
 The $R^2$ values of each slice voxel is mapped on opacity in the 3D-view later on, reducing the occlusion of other values.
 }
  \label{fig:Cube}
\end{figure}
%%
The goal of an overview visualization is to provide a comprehensive view on the data (raw or using descriptive metrics \cite{Bertini}), which is easy to understand.
%% TODO: Cite another paper here!
As described in our previous work \cite{Klemm2014VIS}, correlation values scaled between 0 (no correlation) and 1 (perfect correlation) can be encoded with color in a mosaic plot.
%%
Regression models are more complex, having many associated describing metrics.
%%
For the \emph{Regression Cube} analysis we are interested in the quality-of-fit of the resulting model, which allows to infer about the predictive quality of the independent features included in the model.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the $R^2$ value is the metric allowing for this kind of assessment.
%%
\paragraph{2D (Slice) View.}
%%
Since $R^2$ is scaled between [0,1], it allows for comparison between regression models.
%%
We can apply the same mosaic plot mapping by translating the $R^2$ values to color saturation (Fig.~\ref{fig:Cube} a).
%%
This describes a 2D regression square for dynamic variables $X$ and $Y$ (e.g., $Age \sim X + Y$).
%%

\paragraph{3D (Cube) View.}
%%
Introducing $Z$ creates a 3D \emph{Regression Cube} (Fig.~\ref{fig:Cube} b).
%%
$R^2$ values of each cube entry (\emph{voxel}) is mapped to opacity to reduce the overlap.
%%
The visualization of $R^2$ values derived from different regression cubes (e.g., $Z \sim X + Y$) is misleading, as they can be compared relatively, but not in precise numbers.
%%
Therefore, the $R^2$ results of different regression methods are encoded using different colors (i.e., blue for linear regression and red for logistic regression).
%%
% Adverb immer vors Hauptverb oder Satzende
Thus, the cube can be easily extended using other regression types.
%%
For cubes with a fixed target feature, e.g., $Cancer \sim X + Y + Z$, no such encodings are required and the $z$ dimension can be compared directly.
%%
\\\\
Our goal is to create an overview visualization for a data set, but we also want to incorporate expert knowledge into the visualization by adapting the underlying formulas.
%%
These two approaches do not exclude each other, they rather underline the difference in purpose of the chosen formula.
%%
The different analysis approaches require different starting points using the \emph{Regression Cube}.

\subsection{Analysis Workflow} \label{sec:Workflow}
%%
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow}
 \caption{
 Different workflow types of the analysis using \emph{Regression Cubes} (based on~\cite{Keim}).
 %%
 % The analysis starts with a set of features given by the data set.
 %%
 [\textbf{1.AF}] The workflow starts by declaring a formula regarding to specify a hypothesis, or use a predefined formula for a hypothesis-free analysis.
 %%
 [\textbf{2.SI}] The \emph{Regression Cube} is then visualized.
 %%
 The user has the option to either adjust the formula or to derive details on demand on a specific regression.
 %%
 [\textbf{3.ZF}] Insights into the data yield either an adjustment of the current formula or selecting a difference view.
 %%
 The latter is used to compare regression cubes.
 %%
 [\textbf{4.DD}] Details about features using the 2D mosaic plot representation yields insights and hypothesis about feature relations.
 }
  \label{fig:Workflow}
\end{figure}
%%
%\com{Different Formulas work as different steps in the analysis (Fig.~\ref{fig:Workflow})}
%\com{Workflow for different analysis approaches.}
%\com{0. Select the Variables.}
%\com{1. Select the Formulas}
%\com{a) Analysis without hypothesis about the data set (z ~ x+y)}
%\com{b) Analysis without hypothesis with confounder (z ~ age)}
%\com{c) Analysis with hypothesis about the data set (Var ~ x+y+z)}
%\com{d) Use Difference View}
%\com{d) Clarify using confounder (Var ~ x)}
%\com{3. Find interesting pane.}
%\com{4. Get Details on Demand.}
%%
% Übergang ist recht holprig gewesen
Our Regression Cube models is well-suited for different workflow analysis techniques, based on the Visual Analytics (VA) Mantra of Keim et al. \cite{Keim}:

@@

\textbf{Details on Demand} [\textbf{4.DD}]. Precise information about the individual regression models (coefficients, associated confidence intervals and p-values) can be retrieved based on the data point representatives (e.g. in a hover-modal on a currently selected data point).
\\\\
We use the squared bracket abbreviation for each step to denote the affiliation in the system design section later on.
%%
As seen in Fig~\ref{fig:Workflow}, the workflow is highly iterative.
%%
Observations in the 2D mosaic plot or simply the CFS-based features can trigger new analyses by adjusting the underlying regression formulas.
%%
This can be carried out either to refine the current formula based on observations, or creating a new regression cube for a difference view.
%%
\paragraph{Hypothesis-Free and Hypothesis-Based Analysis.}
%%
Input formulas reflect \emph{hypotheses} about the data.
%%
Using the operators, the dynamic variables and the dataset features, many different assumptions can be expressed.
%%
To support \emph{hypothesis-free} analysis, we provide a default formula:

@@

Hypotheses about the data are easily built up by relating dynamic variables with the regression operators.
%%
Furthermore, static features can be added for each regression formula.
%%
Here are a few examples:

@@

\paragraph{Cube Comparison.}
%%
Regression Cubes can be compared by creating difference views.
%%
One cube (extracted with one formula) acts as reference.
%%
The absolute difference in $R^2$ values towards the second cube is calculated, yielding a difference cube showing only the differences between the two formulas.
%%
For example, it can be utilized for comparing the influence of a single feature towards the complete result (e.g., $Z \sim X + Y$ and $Z \sim X + Y + Income$).
%%
Figure~\ref{fig:Application} (right) shows a comparison cube with subtracted effects of age ($Z \sim Age$).
%%
% \paragraph{Formula Types.}
% %%
% Input formulas reflect \emph{hypotheses} about the data.
% %%
% Using the operators, dynamic variables and dataset features, many different assumptions can be formulated.
% %%
% Here we show just a few examples on how formulas can serve different purposes.
%
% $Z \sim X + Y$ is the default formula.
% %%
% This cube represents all possible combinations of two independent features towards all features in the data set, since we do not know which feature(s) are of interest.
% %%
% Each slice represents a different target feature.
% %%
% It is therefore suitable for a \emph{hypothesis-free} explorative analysis to give a general impression about relationships in the data set.

@@

@@

We designed our system with openness and ease of use in mind.
Using open formats as input interfaces does not restrict the application towards non-epidemiological data sets.

@@

\section{Summary and Conclusion}
%%
We presented a technique for knowledge discovery in cohort study data sets with user-defined target features.
%%
Dimension reduction using the target restricts the analysis to most important features.
%%
\emph{Hypothesis-free} analysis employs default regression models.
%%
Modeling expert knowledge using regression formulas allow for \emph{hypothesis-based} investigation.
%%
A 3D \emph{Regression Cube} allows to assess hotspots in the analysis by abstracting regression models using a quality-of-fit measure.
%%
These can then be analyzed further using the 2D mosaic plot for each cube slice.
%%
Details on demand for each model allow for detailed assessment of regression models.
%%
We successfully applied the approach to find correlations in a hepatic steatosis as well as a breast cancer data set.
%%
The method was well received by our clinical partners, triggering detailed investigations of the findings.

>>>>

% -Dependent & -Based in ‹berschriften groﬂ, sicherheitshalber nochmal Steffi fragen
% oder: Dimension Reduction Depending on the Target Variable
\subsection{Target-Variable-Dependent Dimension Reduction}
% \com{Uli, Can you please proofread this?}
% %%
% The vast majority of features in an epidemiological data have no or a very low relation towards the target feature.
% %%
% Identifying these features und excluding them from the calculation can reduce the number of dimension significantly.
% %%
% The \emph{Correlation based Feature Selection} (CFS) algorithm is very popular in data mining for this task \cite{CFS}.
% %%
% It uses information entropy to select the features which have the most explanatory power towards the target feature.
% %%
% At the same time it tries to reduce the correlation between the independent features.
% %%
In epidemiological studies, manifold recordings lead to an abundance of features and thus a high-dimensional feature space.
%%
In general, many of them exhibit low or none correlation at all towards the target feature.
%%
Identifying irrelevant features and excluding them from the feature space considerably reduces computational costs and yields a comprehensible \emph{Regression Cube} representation.
%%
% nochmal checken wo $F$ eingef¸hrt wird, hier oder im ¸bern‰chsten Satz?
The correlation-based feature selection (CFS) \cite{CFS} aims to find a feature subset that maximizes the \emph{merit value} $M_F$ which is the ratio between the average feature-class and feature-feature dependencies of features in the feature set $F$.
%%
The dependency of a set of features utilizes the entropy-based information gain to measure the explanatory power towards the target feature.
%%
% Ggf. wurde $F$ schon oben eingef¸hrt, TODO
Starting with an empty set of features $F$, the CFS algorithm iteratively adds the feature $f$ to $F$ that leads to the highest new merit-value $M_{F\cup f}$ and halts when no feature is left that would increase the merit.
%%
For example, if the \emph{body weight} has a strong explanatory power towards the target, it is likely that \emph{BMI} or \emph{waist circumference} exhibit similar correlations to the target.
%%
However, they strongly correlate with each other.
%%
The CFS algorithm will select the feature which has the largest explanatory power and discards the other dimensions.

@@

\subsection{Abstracting Regression Results Using $R^2$}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.5in]{figures/cube}
 \caption{
 (left) Overview visualization using a mosaic plot for the formula $Z \sim X + Y$, where $Z$ assumes the feature $Age$.
 %%
 The $R^2$ values extracted from the regression formulas depict the quality-of-fit and are mapped to color saturation (a saturated color shows a strong correlation).
 %%
 (right) Now, $Z$ is set to all features $n$ and yields $n$ mosaic plot visualizations.
 %%
 These represent the slices in our cube.
 %%
 The $R^2$ values of each slice voxel is mapped on opacity in the 3D-view later on, reducing the occlusion of other values.
 }
  \label{fig:Cube}
\end{figure}
%%
The goal of an overview visualization is to provide a comprehensive view on the data (raw or using descriptive metrics \cite{Bertini}), which is easy to understand.
%% TODO: Cite another paper here!
As described in our previous work \cite{Klemm2014VIS}, correlation values scaled between 0 (no correlation) and 1 (perfect correlation) can be encoded with color in a mosaic plot.
%%
Regression models are more complex, having many associated describing metrics.
%%
For the \emph{Regression Cube} analysis we are interested in the quality-of-fit of the resulting model, which allows to infer about the predictive quality of the independent features included in the model.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the $R^2$ value is the metric allowing for this kind of assessment.
%%
\paragraph{2D (Slice) View.}
%%
Since $R^2$ is scaled between [0,1], it allows for comparison between regression models.
%%
We can apply the same mosaic plot mapping by translating the $R^2$ values to color saturation (Fig.~\ref{fig:Cube} a).
%%
This describes a 2D regression square for dynamic variables $X$ and $Y$ (e.g., $Age \sim X + Y$).
%%

\paragraph{3D (Cube) View.}
%%
Introducing $Z$ creates a 3D \emph{Regression Cube} (Fig.~\ref{fig:Cube} b).
%%
$R^2$ values of each cube entry (\emph{voxel}) is mapped to opacity to reduce the overlap.
%%
The visualization of $R^2$ values derived from different regression cubes (e.g., $Z \sim X + Y$) is misleading, as they can be compared relatively, but not in precise numbers.
%%
% e.g. wenn man es w‰hlen kann
% i.e. wenn es so ist
Therefore, the $R^2$ results of different regression methods are encoded using different colors (i.e., blue for linear regression and red for logistic regression).
%%
% Adverb immer vors Hauptverb oder Satzende
Thus, the cube can be easily extended using other regression types.
%%
For cubes with a fixed target feature, e.g., $Cancer \sim X + Y + Z$, no such encodings are required and the $z$ dimension can be compared directly.
%%
\\\\
Our goal is to create an overview visualization for a data set, but we also want to incorporate expert knowledge into the visualization by adapting the underlying formulas.
%%
These two approaches do not exclude each other, they rather underline the difference in purpose of the chosen formula.
%%
The different analysis approaches require different starting points using the \emph{Regression Cube}.

\subsection{Analysis Workflow} \label{sec:Workflow}
%%
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow}
 \caption{
 Different workflow types of the analysis using \emph{Regression Cubes} (based on~\cite{Keim}).
 %%
 % The analysis starts with a set of features given by the data set.
 %%
 [\textbf{1.AF}] The workflow starts by declaring a formula regarding to specify a hypothesis, or use a predefined formula for a hypothesis-free analysis.
 %%
 [\textbf{2.SI}] The \emph{Regression Cube} is then visualized.
 %%
 The user has the option to either adjust the formula or to derive details on demand on a specific regression.
 %%
 [\textbf{3.ZF}] Insights into the data yield either an adjustment of the current formula or selecting a difference view.
 %%
 The latter is used to compare regression cubes.
 %%
 [\textbf{4.DD}] Details about features using the 2D mosaic plot representation yields insights and hypothesis about feature relations.
 }
  \label{fig:Workflow}
\end{figure}
%%
%\com{Different Formulas work as different steps in the analysis (Fig.~\ref{fig:Workflow})}
%\com{Workflow for different analysis approaches.}
%\com{0. Select the Variables.}
%\com{1. Select the Formulas}
%\com{a) Analysis without hypothesis about the data set (z ~ x+y)}
%\com{b) Analysis without hypothesis with confounder (z ~ age)}
%\com{c) Analysis with hypothesis about the data set (Var ~ x+y+z)}
%\com{d) Use Difference View}
%\com{d) Clarify using confounder (Var ~ x)}
%\com{3. Find interesting pane.}
%\com{4. Get Details on Demand.}
%%
% ‹bergang ist recht holprig gewesen
Our Regression Cube models is well-suited for different workflow analysis techniques, based on the Visual Analytics (VA) Mantra of Keim et al. \cite{Keim}:

@@

\textbf{Details on Demand} [\textbf{4.DD}]. Precise information about the individual regression models (coefficients, associated confidence intervals and p-values) can be retrieved based on the data point representatives (e.g. in a hover-modal on a currently selected data point).
\\\\
We use the squared bracket abbreviation for each step to denote the affiliation the system design section later on.
%%
As seen in Fig~\ref{fig:Workflow}, the workflow is highly iterative.
%%
Observations in the 2D mosaic plot or simply the CFS-based features can trigger new analyses by adjusting the underlying regression formulas.
%%
This can be carried out either to refine the current formula based on observations, or creating a new regression cube for a difference view.
%%
\paragraph{Hypothesis-Free and Hypothesis-Based Analysis.}
%%
Input formulas reflect \emph{hypotheses} about the data.
%%
Using the operators, the dynamic variables and the dataset features, many different assumptions can be expressed.
%%
To support \emph{hypothesis-free} analysis, we provide a default formula:

@@

% warum war da ein "either", sollte der Satz etwas anderes ausdr¸cken?
Hypotheses about the data are easily built up by relating dynamic variables with the regression operators.
%%
Furthermore, static features can be added for each regression formula.
%%
Here are a few examples:

@@

\paragraph{Cube Comparison.}
%%
Regression Cubes can be compared by creating difference views.
%%
One cube (extracted with one formula) acts as reference.
%%
The absolute difference in $R^2$ values towards the second cube is calculated, yielding a difference cube showing only the differences between the two formulas.
%%
For example, it can be utilized for comparing the influence of a single feature towards the complete result (e.g., $Z \sim X + Y$ and $Z \sim X + Y + Income$)
%%
% \paragraph{Formula Types.}
% %%
% Input formulas reflect \emph{hypotheses} about the data.
% %%
% Using the operators, dynamic variables and dataset features, many different assumptions can be formulated.
% %%
% Here we show just a few examples on how formulas can serve different purposes.
%
% $Z \sim X + Y$ is the default formula.
% %%
% This cube represents all possible combinations of two independent features towards all features in the data set, since we do not know which feature(s) are of interest.
% %%
% Each slice represents a different target feature.
% %%
% It is therefore suitable for a \emph{hypothesis-free} explorative analysis to give a general impression about relationships in the data set.

@@

%% scientific english: kurz und langweilig ;)
We designed our system to be freely accessible and easy to use.
With open formats as input interfaces, the application can be extended to non-epidemiological data sets.

@@

@@

\section{Summary and Conclusion}
%%
We presented a technique for knowledge discovery in cohort study data sets with user-defined target features.
%%
Dimension reduction using the target restricts the analysis to most important features.
%%
\emph{Hypothesis-free} analysis employs default regression models.
%%
Model expert knowledge using regression formulas allow for \emph{hypothesis-based} investigation.
%%
A 3D \emph{Regression Cube} allows to assess hotspots in the analysis by abstracting regression models using a quality-of-fit measure.
%%
These can then be analyzed further using the 2D mosaic plot for each cube slice.
%%
Details on demand for each model allow for detailed assessment of regression models.
%%
We successfully applied the approach to find correlations in a hepatic steatosis as well as a breast cancer data set.
%%
The method was well received by our clinical partners, triggering detailed investigations of the findings.
